---
title: "RT-PCR data analysis - hypothesis testing on the existence of treatment effect"
author: "Byungwon Kim, Johan Lim, Sungkyu Jung, Woncheol Jang"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: vignette
---


This is an R code with an example provided for analyzing an RT-PCR experiment data or other data with similar characteristics summarized as follows:

1. small sample size (usually smaller than 20 including replicates)
2. repeatedly measured observations or technical replicates from the same subject 
3. two groups(populations) are compared - treatment group versus control group

First of all, installation of the follwing R packages is required for the analysis.

```{r setup, include = TRUE, echo = TRUE}
knitr::opts_chunk$set(echo = TRUE)
# Please run these code if you don't have these packages installed
#install.packages("gee")
#install.packages("gtools")
#install.packages("tidyverse")
#install.packages("geesmv")
```

If have them already or after the installation, please load them on your system by running the following code.

* library(gee)
* library(gtools)
* library(tidyverse)
* library(geesmv)

```{r library setup, echo = FALSE, include = FALSE}
library(gee)
library(gtools)
library(tidyverse)
library(geesmv)
```

Now, we will define a few functions which will help us to get the results of hypothesis testing from the methods dealt in the manuscript.
Details are provided in every function for users who are interested in learning the R code.
If want to get just the results, please run the following code just once and skip the details.

* We omit the code on this output
* The original R markdown file contains the functions

```{r functions, echo = FALSE, include = FALSE}
Data.rearrangement <- function(data, rearrange.id, group, subject){
  # Function for data rearrangement
  # Data must have the variable for index of subject
  # Input:
  #       data - raw data frame 
  #       rearrange.id - index for rearrangement
  #       group - name of group variable
  #       subject - name of subject id variable
  # Output:
  #       data.re - subject id variable is changed for rearrangement
  group.id <- data[,group]
  subject.id <- data[,subject]
  data.order <- data[order(group.id, subject.id),]
  data.re <- data.order[data.order[,subject] == rearrange.id[1],]
  for (j in 2:length(rearrange.id)){
    data.temp <- data.order[data.order[,subject] == rearrange.id[j],]
    data.re <- rbind(data.re, data.temp)
  }
  data.re[,group] <- sort(group.id) 
  return(data.re)
}

t.ind.stat = function(data, response, group){
  # Calculates the test statistic for two-sample t-test
  # (M1) in the manuscript
  # Treats the replicates as independent observations
  # Test is fixed for the alternative "greater"
  # which means that the treatment effect is positive
  # Input:
  #         data - usual data frame with 3 variables (response, group, subject)
  #         response - name of response variable
  #         group - name of group variable
  #                 1: treated as the treatment group
  #                 2: treated as the control group
  # Output:
  #         test.stat - test statistic 

  test.stat <- t.test(data[,response] ~ data[,group], equal.var = T)$statistic  
  return(test.stat)
}

wilcoxon.stat = function(data, response, group, alternative){
  # Calculates the test statistic for Wilcoxon's rank sum test
  # (M2) in the manuscript
  # Treats the replicates as independent observations
  # Input:
  #         data - usual data frame with 3 variables (response, group, subject)
  #         response - name of response variable
  #         group - name of group variable
  #                 1: treated as the treatment group
  #                 2: treated as the control group
  # Output:
  #         test.stat - test statistic 
  
  test.stat <- wilcox.test(data[,response] ~ data[,group])$statistic
  return(test.stat)
}

t.avg.stat = function(data, response, subject, group){
  # Calculates the test statistic for two-sample t-test
  # (M3) in the manuscript
  # Take the average of replicates from the same subject and
  # treats those averages as independent sample
  # Input:
  #         data - usual data frame with 3 variables (response, group, subject)
  #         response - name of response variable
  #         subject - name of subject variable
  #         group - name of group variable
  #                 1: treated as the treatment group
  #                 2: treated as the control group
  # Output:
  #         test.stat - test statistic 
  
  id <- data[,subject]
  unique.id <- unique(id)
  data.avg <- data.frame(res = rep(0, length(unique.id)),
                         gr = rep(0, length(unique.id)))
  names(data.avg) <- c(response, group)
  for (j in 1:length(unique.id)){
    data.avg[j,] <- c(mean(data[data[,subject] == unique.id[j], response]), mean(data[data[,subject] == unique.id[j], group]))
  }
  test.stat <- t.test(data.avg[,response] ~ data.avg[,group], equal.var = T)$statistic
  return(test.stat)
}

GEE.stat = function(data, response, subject, group){
  # Calculates the test statistic for GEE Wald test
  # (M4) in the manuscript
  # Input:
  #         data - usual data frame with 3 variables (response, group, subject)
  #         response - name of response variable
  #         subject - name of subject variable
  #         group - name of group variable
  #                 1: treated as the treatment group
  #                 2: treated as the control group
  # Output:
  #         test.stat - test statistic 
  
  gee.fit <- gee(data[,response] ~ data[,group],
                 id = data[,subject],
                 family = gaussian,
                 corstr = "exchangeable") %>% summary()
  test.stat <- gee.fit$coefficients[2,5]
  if (mean( data[data[,group] == 1, response] ) - mean( data[data[,group] == 2, response] ) > 0){
    test.stat <- abs(test.stat)
  }else{
    test.stat <- -abs(test.stat)
  }
  
  return(test.stat)
}

Permutation.pvalue <- function(data, response, subject, group, alternative){
  # Calculates the permutation p-values for (M1) -- (M4) in the manuscript
  # The rule for permutation is also given in the manuscript
  # Input:
  #         data - usual data frame with 3 variables (response, group, subject)
  #         response - name of response variable
  #         subject - name of subject variable
  #         group - name of group variable
  #         alternative - "two.sided": treatment effect != control effect
  #                       "greater"  : treatment effect > control effect
  #                       "less"     : treatment effect < control effect
  # Output:
  #         p.values - permutation p-values for all (M1)--(M4) test statistics
  
  n <- length(unique(data[,subject]))
  n1 <- length(unique(data[data[,group] == unique(data[,group])[1],subject]))
  n2 <- length(unique(data[data[,group] == unique(data[,group])[2],subject]))
  # permutation index
  perm.ids = gtools::combinations(n, n1)
  
  # Our empirical rule says, if the sample size gives more than 2000 permutation combinations,
  # then randomly chosen 2000 combinations would be enough for p-value calculation.
  # Recall, permutation approach is appropriate for small sample size case.
  if (nrow(perm.ids) > 2000){
    perm.ids <- perm.ids[sample(nrow(perm.ids), 2000, replace = FALSE),]
  }
  
  allperms_ids = apply(perm.ids, 1,  function(x){
    c(x, (1:n)[-x])
  }) %>% t()
  
  test.stat = c( t.ind.stat(data, response, group),
                 wilcoxon.stat(data, response, group),
                 t.avg.stat(data, response, subject, group),
                 GEE.stat(data, response, subject, group) )
  
  names(test.stat) = c("t_ind", "Wilcoxon", "t_avg", "gee.wald")
  
  # permutation p-value
  perm.stat <- data.frame(t_ind = rep(0, nrow(allperms_ids)),
                          Wilcoxon = rep(0, nrow(allperms_ids)),
                          t_avg = rep(0, nrow(allperms_ids)),
                          gee.wald = rep(0, nrow(allperms_ids)))
  perm.stat[1,] <- test.stat
  for(j in 2:nrow(allperms_ids)){
    data.re <- Data.rearrangement(data, rearrange.id = allperms_ids[j,], group, subject)
    perm.stat[j,] <- c( t.ind.stat(data.re, response, group),
                        wilcoxon.stat(data.re, response, group),
                        t.avg.stat(data.re, response, subject, group),
                        GEE.stat(data.re, response, subject, group) )
  }
  
  pvalues = vector(length = length(test.stat))
  for (j in 1:length(test.stat)){
    if (alternative == "two.sided"){
      pvalues[j] = 2*mean(perm.stat[,j] >= abs(test.stat[j])) 
    }else if (alternative == "greater"){
      pvalues[j] = mean(perm.stat[,j] >= test.stat[j]) 
    }else if (alternative == "less"){
      pvalues[j] = mean(perm.stat[,j] <= test.stat[j]) 
    }else{
      print("Not an appropriate alternative hypothesis!!")
    }
  }
  return(pvalues)
}
```

From now on, we will give a real data example. Users will be able to get the results for their own data by only importing the data they want to analyze.
Here, we use the data in Fu et al.(2006). 

```{r data example, echo = FALSE, include = FALSE}
Fu.data <- data.frame(deltact = c(3.70, 3.36, 2.02, 2.47, 3.22, 3.28, 2.68, 1.80, 2.53, 2.58, 2.28, 2.48),
                      sample = c(1, 1, 2, 2, 3, 3, 4, 4, 5, 5, 6, 6),
                      treatment = c(1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2))

# Permutation p-values
p.values.two.sided <- Permutation.pvalue(Fu.data, response = "deltact", subject = "sample", group = "treatment", alternative = "two.sided")
p.values.one.sided <- Permutation.pvalue(Fu.data, response = "deltact", subject = "sample", group = "treatment", alternative = "greater")

```

```{r data table, echo = TRUE}
print(Fu.data)
```

To get the permutation p-values, one can run the following code.

* p.values <- Permutation.pvalue(Fu.data, response = "deltact", subject = "sample", group = "treatment")

It provides the below p-values.

```{r example result}
print(p.values.two.sided)
print(p.values.one.sided)
```

